---
title: "Tidy Netherlands"
author: "Amber Kalb"
date: "3/18/2017"
output: html_document
---



```{r, echo=FALSE}
library(dplyr)
library(reshape2)
library(ggplot2) 

new_neth$id <- rownames(new_neth) 
melt(new_neth)

new_neth$variable <- NULL
head(new_neth)
```


```{r, echo=FALSE}

New_Neth2<- new_neth

New_Neth2 <- New_Neth2[c("id", "Year", "Region", "Party", "Alignment", "Votes N", "Votes %")]

write.csv(New_Neth2, file="New_Neth2.csv")


```
## Exploratory Plots

```{r}
New_Neth2%>%
  filter(Region == "Netherlands") %>%
  group_by(Year, Party, Region) %>%
  ggplot(aes(x = Year, y = `Votes %`, color = Party)) +
  geom_line()+
  labs(title = "Party competition in the Netherlands",
       x = NULL, y = "Percentage of vote") +
  theme_minimal() +
  theme(legend.position = "bottom")
```
#So in this visualization we have two parties in 2012 elections really taking off: Pvda (centre left) and the VVD(centre right)

```{r, echo=FALSE}

Neth_Align <- New_Neth2%>%
  filter(Region == "Netherlands") %>%
  group_by(Year, Region, Alignment) %>% 
  summarise(Votes = sum(`Votes N`, na.rm = TRUE)) %>% 
  group_by(Year) %>% 
  mutate(Total_Votes = sum(Votes)) %>% 
  mutate(Percent_votes = Votes / Total_Votes)
  write.csv(Neth_Align, file="Neth_Align.csv")

  ggplot(Neth_Align, aes(x = Year, y = Percent_votes, color = Alignment)) +
  geom_line()+ geom_point()+
  labs(title = "Party competition in the Netherlands",
       x = "Year", y = "Percentage of vote") +
  theme_minimal() 
 
```
# So the centre is losing voters, they are polarizing to centre left or right, meanwhile the alt-factions are moving towards the center. Looking at the Party graph you can see that the shift is in relation to the VVD and the Pvda parties.

```{r}
New_Neth2%>%
  group_by(Year, Party, Region) %>%
  ggplot(aes(x = Year, y = `Votes %`, color = Party)) +
  geom_line()+
  labs(title = "Party competition in the Netherlands",
       x = NULL, y = "Percentage of vote") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  facet_wrap(~ Region, ncol = 5)
```

```{r}
New_Neth2%>%
  filter(Party == "PVV") %>%
  group_by(Year, Party, Region) %>%
  ggplot(aes(x = Year, y = `Votes %`, color = Region)) +
  geom_line()+
  labs(title = "Votes for the PVV by Region",
       x = "Year", y = "Percentage of Vote", legend = "Region") +
  theme_minimal() +
  theme(legend.position = "bottom") 
  
```

```{r}
ggplot(Neth_Align,aes(x= Year, y = Percent_votes, group = Alignment, fill=Alignment)) + geom_area(position="identity",alpha=.5)
```
# Embedded Plots

```{r pressure, echo=FALSE}
p <- ggplot(Neth_Align, aes(x= Year, y= Percent_votes,fill= Alignment)) 
p + geom_bar(stat="identity")
```
```{r}


Stacked_Align<- ggplot(Neth_Align, aes(x= Year, y= Percent_votes)) + 
  geom_area(aes(fill= Alignment, group = Alignment), position='stack') + scale_fill_brewer()

Stacked_Align  
  
```

```{r}
Neth_Party<- New_Neth2%>%
  filter(Region == "Netherlands") %>%
  group_by(Year, Party) %>% 
  summarise(Votes = sum(`Votes N`, na.rm = TRUE)) %>% 
  group_by(Year) %>% 
  mutate(Total_Votes = sum(Votes)) %>% 
  mutate(Percent_votes = Votes / Total_Votes) 
  
 write.csv(Neth_Party, file="Neth_Party.csv")

 Neth_Party %>% 
   group_by(Year, Party) %>% 
ggplot(aes(x= Year, y= Percent_votes)) + 
  geom_area(aes(fill= Party, group = Party), position='stack') + scale_fill_brewer() %>% View()

```

#So I really like this visualization due to the facr that you can really see were 



```{r, echo=FALSE,   warning= FALSE}
New_Neth2%>%
  filter(Party == "PVV") %>%
  group_by(Year, Party, Region) %>%
  ggplot(aes(x = Year, y = `Votes %`, color = Region)) +
  geom_line()+
  labs(title = "Votes for the PVV by Region",
       x = "Year", y = "Percentage of Vote", legend = "Region") +
  theme_minimal() +
  theme(legend.position = "bottom") 
  
```





